\chapter{Acknowledgements}

Thank you, Fr\'ed\'eric, for introducing me to your way of thinking about physics. Having a clear mathematical foundation is fundamental for asking the right questions later on---but the mathematics does never become an end in itself. Together with the whole Constructive Gravity Group, we embraced this principle and delivered interesting results using innovative methods, despite facing challenging conditions. I especially enjoyed the deep dive with Tobias into the covariant approach and the computer-aided solution methods. What we two achieved in relatively short time makes me proud.

The students at the chair for Theoretical Physics III should not remain unmentioned. It was always nice having them around, be it for stimulating discussions, coffee breaks, or the occasional game of table football. I also had the privilege of getting to know a lot of wonderful coworkers during my former employment at Interflex Datensysteme GmbH and now at TNG Technology Consulting GmbH. By now, I see it as an advantage not to be solely occupied with academic work---one has to leave his comfort zone, which leads to mental flexibility.

I am very grateful to Prof.~Gerd Leuchs from the Max Planck Institute for the Science of Light for supervising my PhD.

However, I am above all indebted to my friends and family for their unconditional support throughout this challenging time. Without my parents, nothing of this would have been possible. Without you, Jennifer, it could not have been this enjoyable.

